---
title: "Bayesian Logistic Regression - Application in Probability Prediction of disease (Diabetes) "
subtitle: "CapStone Project_2025"
author: "Namita Mishra, Autumn Wilcox, Ecil Teodoro (Advisor: Dr. Ashraf Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

------------------------------------------------------------------------

## Introduction

Diabetes mellitus (DM) represents a major public health concern and is
closely associated with factors such as obesity, age, race, and gender.
Identifying these associated risk factors is essential for developing
targeted and effective intervention strategies @DAngelo2025. **Logistic
Regression** that estimates the association between risk factors and
binary outcomes (presence or absence of diabetes) is a standard
analytical approach but, is insufficient in analyzing the complexity of
healthcare data (DNA sequences, imaging, patient-reported outcomes,
electronic health records (EHRs), longitudinal health measurements,
diagnoses, and treatments. @Zeger2020. Classical maximum likelihood
estimation (MLE) yields unstable results in small samples with missing
data, or quasi- and complete separation.

The Bayesian hierarchical model with Markov Chain Monte Carlo (MCMC) on
multivariate longitudinal healthcare data with two levels: (1) repeated
measures over time within individuals and (2) individuals nested within
a population, by integrating prior knowledge and by added exogenous
covariates (e.g., age, clinical history), and endogenous covariates
(e.g., current treatment), yield posterior distributions and marginal
distributions from MCMC estimation of parameters provide risk prediction
(pneumonia, prostate cancer, and mental disorders). The model's
limitation is its parametric nature.

In Bayesian inference @Chatzimichail2023, comparisons between parametric
models (with fixed sets of parameters) and non-parametric models (which
make no prior assumptions about the distribution) applied to the
National Health and Nutrition Examination Survey demonstrated the use of
posterior probabilities to classify diseases. In contrast, conventional
approaches that rely on fixed clinical thresholds and dichotomous
classifications often fail to capture the complex relationships between
diagnostic tests and disease prevalence. These traditional methods tend
to oversimplify the underlying heterogeneity across populations and
perform poorly when dealing with skewed, bimodal, or multimodal data
distributions.

Bayesian nonparametric modeling is flexible, adaptable, and robust
alternative to traditional parametric approaches capturing complex and
multimodal data patterns by integration of priors with multiple
diagnostic tests enhances diagnostic accuracy and precision. However,
the approach can be limited by restricted access to scholarly resources
and excessive reliance on prior information. Combined with other
statistical and computational techniques, Bayesian nonparametric models
can further strengthen diagnostic performance addressing systemic bias,
incomplete data, and non-representative or non-normal datasets
@Chatzimichail2023

The Bayesian framework described by @VandeSchoot2021 highlights the
critical role of priors, data modeling, inference, model checking,
posterior sampling, variational inference, and variable selection in
applications across social sciences, ecology, genetics, and medicine.
Variable selection is important because multicollinearity, limited
sampling, and overfitting can lead to weak predictive performance and
reduced interpretability. Based on the degree of uncertainty in
hyperparameters, priors are classified as informative, weakly
informative, or diffuse (larger variances indicating greater
uncertainty).

Prior elicitation may be based on expert opinion, generic knowledge, or
data-driven using maximum likelihood estimates or sample statistics.
Prior sensitivity analysis assesses the alignment between priors and the
likelihood. Prior improve estimation via data-informed shrinkage,
regularization, and influence on algorithms by focusing on high-density
regions. Specifying prior information strengthens inference by
integrating observed data with uncertain parameters in small or less
informative samples, generating a range of possible values through the
likelihood function. The Markov Chain Monte Carlo (MCMC) simulation
algorithm, ( **brms** and **blavaan** in R), allows empirical estimation
of posterior distributions. Spatial and temporal Bayesian models analyze
large-scale cancer genomics, identify novel molecular interactions,
detection of mutational signatures, genomic-based patient stratification
for clinical trials, and insights into cancer evolution, but ahve
limited application because of temporal autocorrelation and the inherent
subjectivity in prior elicitation.

Bayesian normal linear regression, has been applied in metrology for
instrument calibration using parametric (conjugate) models with a
Normal–Inverse-Gamma (NIG) prior @Klauenberg2015 as in Gaussian
regression, errors are independent and identically distributed with
unknown variance, the relationship between X and Y is statistical and
due to the associated noise and model uncertainty, can not be treated as
a direct measurement function. The Bayesian method accounts for
uncertainty, treats both observables (data) and unobservables
(parameters and auxiliary variables) as random and assigns probability
distributions to unobservables. Incorporation of prior, enhances
interpretability and robustness. The NIG distribution (conjugate prior)
enables the use of vague or non-informative priors, and hierarchical
priors further add flexibility by modeling uncertainty across multiple
levels, and capture complex data nuances.

**Bayesian Hierarchical / meta-analytic linear regression** model
augments data by incorporating both exchangeable and unexchangeable
information on parameters, addressing issues associated with multiple
testing providing low statistical power when separate significance tests
are conducted across studies with different predictors. Adding priors
from meta-analysis in Bayesian Hierarchical regression address the
challenge of small sample size with limited availability of previous
articles, and resolve limitation of univariate analyses, and the
relationship issues among multiple regression parameters within a study.
(1) Exchangeable Priors are when the current data and previous studies
share the same set of predictors, and (2) Unexchangeable when the
predictors differ. The hierarchical unexchangeable model is applicable
in studying differences in studies, enabling explicit testing of the
exchangeability assumption. Approach is limited due to the correlation
between identical set of predictors. @DeLeeuw2012a

**A sequential clinical reasoning model**\*\* @Liu2013 demonstrated
screening adults (20–79 years, Taiwan) while addressing the limited
availability of molecular information and provided an alternative method
leveraging routinely collected biological markers classifying diseases.
Sequential adding of predictors in three models: (1) demographic
features (basic model), (2) six metabolic components (metabolic score
model), and (3) conventional risk factors (enhanced model),
incorporating priors, and emulating a clinician’s evaluation process,
the model assumes normally distributed regression coefficients, accounts
for uncertainty in clinical weights, and averages credible intervals for
predicted risk estimates. by incorporating patient background,
individual characteristics, in the enhanced model, significantly
contributed to the baseline risk estimation by capturing ecological
heterogeneity. The model limitation is interactions between predictors
and the external cross-validation issue.

**Bayesian multiple imputation with logistic regression** addresses
missing data in clinical research **@Austin2021** by classifying and
analyzing missing values based on (i) patients' refusal (ii) loss to
follow-up, (iii) investigator or mechanical errors, or (iv) physicians'
choice on ordering investigations. Missing at random (MAR), missing not
at random (MNAR), or missing completely at random (MCAR) are crucial in
data analysis. Multiple imputation (MI) (R, SAS, or Stata) provides
plausible values while simultaneously conducting identical statistical
analyses across them, creating multiple completed datasets, and the
pooled results classify patients' health status and mortality rates.

**Aims**

The study aims to apply Bayesian logistic regression to predict diabetes
status and to evaluate the associations between body mass index (BMI),
age (≥20 years), gender, and race as predictors, using a retrospective
dataset from the 2013–2014 NHANES survey. NHANES employs a complex
sampling design, including stratification, clustering, and oversampling
of specific population subgroups, rather than uniform random sampling. A
Bayesian analytical approach is used to address challenges posed by
dataset anomalies such as missing data, complete case analysis, and
separation that limit the efficiency and reliability of traditional
logistic regression in predicting health outcomes.

## Method and Data Preparation

**Statistical Tool**: we use R, R packages and libraries to import data,
perform data wrangling and analysis.

**Data source:** NHANES 2-year data is a cross-sectional weighted data
(2013-2014 year) @CenterforHealthStatistics1999. Three datasets
(demographics, exam, questionnaire) imported (Haven package to coverted
.XPT files in R to dataframe (df)) and after selecting variables of
interest a merged dataset was created.

**Modeling** - We conduct bayesian logisitc regression to estimate the
association between BMI, age, sex, and race/ethnicity and predict
doctor-diagnosed diabetes (`DIQ010`).

```{r}
#| label: Libraries
#| echo: true


# loading packages 
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("nhanesA")
library ("nhanesA")    

library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(Hmisc)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(classpackage)
library(janitor)
install.packages("gt")   
library(gt)
library(survey)
library(DataExplorer)
library(logistf)


```

**Method**

**Data pre-processing and cleaning** Subsets are created from the
original weighted datasets (demographics, exam, questionnaire) with
selected variables are merged using ID to create a single dataframe.

1.  Response Variable: Binary, **type 2 / diagnosed diabetes**
    (excluding gestational diabetes) -“Doctor told you have diabetes?”
    DIQ010 combined with `DIQ050` a **secondary variable** describing
    **treatment status (insulin use) to exclude those cases**
2.  Predictor Variables (Body Mass Index, factor, 4 levels are analyzed
    after standardization).\
    o Underweight (\<5th percentile)\
    o Normal (5th–\<85th)\
    o Overweight (85th–\<95th) o Obese (≥95th percentile)\
    o Missing We kept it as it is as categorization provides clinically
    interpretable groups
3.  Covariates:
    1.  Gender (factor, 2 levels): Male: Female
    2.  Ethnicity (factor, 5 levels): Mexican American, Non-Hispanic,
        White Non-Hispanic, Black Other Hispanic, Other Race - Including
        multi-racial
    3.  Age (number, continuous)

```{r message=FALSE, warning=FALSE}
#| label: nhanes tables and merged data
#| include: false

            

library ("nhanesA")     
 # imported datasets 
                       nhanesTables('EXAM', 2013)
                       nhanesTables('QUESTIONNAIRE', 2013)
                       nhanesTables('DEMOGRAPHICS', 2013)
         
                                     
 # codebook for variable details

nhanesCodebook("DEMO_H",'RIDRETH1')
nhanesCodebook("DEMO_H",'RIAGENDR')
nhanesCodebook("DEMO_H",'RIDAGEYR')
nhanesCodebook("DIQ_H","DIQ010, DIQ050")
nhanesCodebook("BMX_H",'BMDBMIC')

  #  .xpt files read ( 2013–2014)                      
                      bmx_h <- nhanes("BMX_H")         #Exam
                      demo_h <- nhanes("DEMO_H")       #Demo
                      diq_h <- nhanes("DIQ_H")         #diabetes


# variables of interest

library(dplyr)

exam_sub <- bmx_h %>% 
  select(SEQN, BMDBMIC) %>%
  rename(
    ID = SEQN,
    BMI = BMDBMIC
  )


need_demo <- c("SEQN","RIDAGEYR","RIAGENDR","RIDRETH1","SDMVPSU","SDMVSTRA","WTMEC2YR")
stopifnot(all(c("SEQN","BMXBMI") %in% names(bmx_h)))
stopifnot(all(need_demo %in% names(demo_h)))
if (!("DIQ010" %in% names(diq_h))) {
  stop("DIQ010 is not in DIQ_H. Check the cycle name 'DIQ_H' and nhanesA version.")
}


      # ---- Select only needed variables ----
exam_sub <- bmx_h  %>% select(SEQN, BMXBMI)
demo_sub <- demo_h %>% select(all_of(need_demo))
diq_sub  <- diq_h  %>% select(SEQN, DIQ010, dplyr::any_of("DIQ050"))

# merged dataframe

merged_data <- demo_sub %>%
  left_join(exam_sub, by = "SEQN") %>%
  left_join(diq_sub,  by = "SEQN")

names(merged_data)
saveRDS(merged_data, "data/nhanes2013_2014_prepared.rds")


```

Merged dataset created, cleaned and visualized for any anomalies and ata
Exploration.

```{r}
#| label: data structure and missing data
#| echo: true

# weighted means of each variable                       
str(merged_data)
plot_str(merged_data)
introduce(merged_data)

p1 <- plot_intro(merged_data, title="Figure 1 (Merged dataset). Structure of variables and missing observations.")
p2 <- plot_missing(merged_data, title="Figure 2(Merged dataset). Breakdown of missing observations.")


# Save it as a PNG file
ggsave("Figure1_MergedDataset.png", plot = p1, width = 8, height = 6, dpi = 300)


ggsave("Figure2_MergedDataset.png", plot = p2, width = 8, height = 6, dpi = 300)

```

```{r message=FALSE, warning=FALSE}
#| label: merged dataframe and character handling
#| echo: true


# print(glimpse(merged_data))
print(table(merged_data$BMDBMIC, useNA = "ifany"))
print(table(merged_data$DIQ010,  useNA = "ifany"))

# ---- Coercion helpers (handle labelled/character) ----
to_num <- function(x) {
  if (is.numeric(x)) return(x)
  xc <- as.character(x)
  n <- suppressWarnings(readr::parse_number(xc))
  if (mean(is.na(n)) > 0.80) {
    xlow <- tolower(trimws(xc))
    n <- dplyr::case_when(
      xlow %in% c("1","yes","yes, told") ~ 1,
      xlow %in% c("2","no","no, not told") ~ 2,
      xlow %in% c("3","borderline") ~ 3,
      xlow %in% c("7","refused") ~ 7,
      xlow %in% c("9","don't know","dont know","unknown") ~ 9,
      TRUE ~ NA_real_
    )
  }
  as.numeric(n)
}

merged_data <- merged_data %>%
  mutate(
    DIQ010   = to_num(DIQ010),
    DIQ050   = to_num(if (!"DIQ050" %in% names(.)) NA_real_ else DIQ050),
    BMXBMI   = suppressWarnings(as.numeric(BMXBMI)),
    RIDAGEYR = suppressWarnings(as.numeric(RIDAGEYR)),
    RIAGENDR = suppressWarnings(as.numeric(RIAGENDR)),
    RIDRETH1 = suppressWarnings(as.numeric(RIDRETH1)),
    SDMVPSU  = suppressWarnings(as.numeric(SDMVPSU)),
    SDMVSTRA = suppressWarnings(as.numeric(SDMVSTRA)),
    WTMEC2YR = suppressWarnings(as.numeric(WTMEC2YR))
  )

# ---- Diagnostics BEFORE save ----
cat("DIQ010 counts BEFORE save:\n")
print(table(merged_data$DIQ010, useNA = "ifany"))
cat("Count with DIQ010 in {1,2}:", sum(merged_data$DIQ010 %in% c(1,2), na.rm = TRUE), "\n")

# ---- Save to file for reuse ----
dir.create("data", showWarnings = FALSE)
# ---- Save ----
dir.create("data", showWarnings = FALSE, recursive = TRUE)
saveRDS(merged_data, "data/merged_2013_2014.rds")
message("Saved: data/merged_2013_2014.rds")


```

**EDA**

-   Used library(survey) to get weighted means and sd of the variables.
    The BMI and age were standardized.
-   Age was recoded into different variabless, including only \>20 years
    in the analysis.
-   BMI is recoded and categorized
    as-"18.5,18.5–\<25,25–\<30,30–\<35,35–\<40,≥40 years).
-   Ethnicity is recoded as "Mexican American" = "1", "Other Hispanic" =
    "2", "NH White" = "3", "NH Black" = "4", "Other/Multi" = "5"
-   Since special codes are not random, cannot be dropped; the
    informative missingness if ignored (MAR or MNAR) could introduce
    bias. We transformed special codes (3,7,) to NA and included all NAs
    in the analysis. Visulaization of missing data presented below.
-   A final analytic dataset was created ('adult') with "NH White" and
    "Male" as the reference group

```{r}
#| label: Basic Exploration (adults)
## 
# ---------------- Basic Exploration (adults) ----------------

# Keep adults only and build analysis variables
adult <- merged_data %>%
  dplyr::filter(RIDAGEYR >= 20) %>%
  dplyr::transmute(
    # --- keep survey design variables so svydesign() can see them ---
    SDMVPSU, SDMVSTRA, WTMEC2YR,

    # --- outcome: DIQ010 (1 yes, 2 no; 3/7/9 -> NA) ---
    diabetes_dx = dplyr::case_when(
      DIQ010 == 1 ~ 1,
      DIQ010 == 2 ~ 0,
      DIQ010 %in% c(3, 7, 9) ~ NA_real_,
      TRUE ~ NA_real_
    ),

    # --- predictors (raw) ---
    bmi  = BMXBMI,
    age  = RIDAGEYR,

    # sex (1=Male, 2=Female)
    sex  = forcats::fct_recode(factor(RIAGENDR), Male = "1", Female = "2"),

    # race (5-level)
    race = forcats::fct_recode(
      factor(RIDRETH1),
      "Mexican American" = "1",
      "Other Hispanic"   = "2",
      "NH White"         = "3",
      "NH Black"         = "4",
      "Other/Multi"      = "5"
    ),

    # keep DIQ050 so we can safely reference it (may be absent/NA in some rows)
    
    DIQ050 = DIQ050
  ) %>%
  # standardize continuous predictors
  dplyr::mutate(
    age_c = as.numeric(scale(age)),
    bmi_c = as.numeric(scale(bmi)),
    bmi_cat = cut(
      bmi,
      breaks = c(-Inf, 18.5, 25, 30, 35, 40, Inf),
      labels = c("<18.5","18.5–<25","25–<30","30–<35","35–<40","≥40"),
      right = FALSE
    )
  ) %>%
  # adjust outcome: if female & DIQ050==1 ("only when pregnant"), set to 0 (not diabetes)
  dplyr::mutate(
    diabetes_dx = ifelse(sex == "Female" & !is.na(DIQ050) & DIQ050 == 1, 0, diabetes_dx)
  )

# Make NH White the reference level for race (clearer interpretation)
adult <- adult %>%
  dplyr::mutate(
    race = forcats::fct_relevel(race, "NH White")
  )

# --- sanity checks ---
cat("Adults n =", nrow(adult), "\n")



```

```{r}
#| label: Data exploration
# data exploration

print(table(adult$diabetes_dx, useNA = "ifany"))
print(table(adult$sex, useNA = "ifany"))
print(table(adult$race, useNA = "ifany"))

if (sum(!is.na(adult$diabetes_dx)) == 0) {
  stop("Too few non-missing outcomes for modeling (n = 0). Check DIQ010 upstream.")
}

# (optional plots omitted for brevity)

# save for downstream
if (!dir.exists("data")) dir.create("data", recursive = TRUE)
saveRDS(adult, "data/adult_cleaned_2013_2014.rds")

```

```{r}
#| label: survey design
# survey design
# ---------------- Survey Design ----------------
# Use exam weights because BMI (BMXBMI) is an MEC variable
nhanes_design_adult <- survey::svydesign(
  id = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC2YR,
  nest = TRUE,
  data = adult
)

# quick weighted checks
survey::svymean(~age, nhanes_design_adult, na.rm = TRUE)
survey::svymean(~diabetes_dx, nhanes_design_adult, na.rm = TRUE)


```

\*\* Statistical Methods\*\*

## Modeling

**Multiple Logistic regression** on survey weighted dataset

-We conducted frequentist method **Multiple Logistic regression** on a
survey-weighted dataset, for complete case analysis and data exploration

**Multivariate Imputation by Chained Equations (MICE)** - We then
conducted MICE to manage missiging data - Considering the small sample
size, Multivariate Imputation by Chained Equations (MICE) was conducted
as an alternative to the Bayesian Approach @JSSv045i03

-   Multiple imputation (MI) is an alternative analytic approach for
    small dataset with missingness.

-   Flatness of the density, heavy tails, non-zero peakedness, skewness
    and multimodality do not hamper the good performance of multiple
    imputation for the mean structure in samples n \> 400 even for high
    percentages (75%) of missing data in one variable
    \@@van2012flexible.

-   Multiple Imputation (MI) can be performed using mice package in R
    and adds sampling variability to the imputations.

-   Iterative MICE imputes missing values of one variable at a time,
    using regression models based on the other variables in the dataset.

-   In the chain process, each imputed variable become a predictor for
    the subsequent imputation, and the entire process is repeated
    multiple times to create several complete datasets, each reflecting
    different possibilities for the missing data.

**Bayesian Logistic Regression**

Bayesian statistics is about updating beliefs with evidence:

Posterior ∝ Likelihood × Prior

-   Prior (p(θ)): Your initial belief about a parameter before seeing
    the data.

-   Likelihood (p(y\|θ)): How probable the observed data are given the
    parameters. This is derived from the model (e.g., logistic
    regression likelihood).

-   Posterior (p(θ\|y)): Your updated belief about the parameter after
    seeing the data.

-   We selected Bayesian Logistic Regression since our study response
    variable is a binary outcome (Diabetes:yes/no)

-   Bayesian Logistic Regression is based on binomial probability Bayes'
    rules, and predicts probability of disease outcome

-   Bayes analyzes linear relation between the predictor (Age, Race,
    BMI, Gender) and outcome response variable (Diabetes).

-   it considers that predictors and response variables are independent.

-   Regression a of a discrete variable (0 or 1) is a Bernoulli
    probability model that classifies categorical response variables -
    predicting Diabetes.

-   Logit link provides probabilities for the response variable.

-   We use Weakly informative priors Normal (0, 2.5) for logistic
    regression coefficients and intercept, Normal(0, 10), allows a wide
    range of baseline log-odds and helps with convergence and avoids
    extreme estimates. Good default for most applications in social,
    health, or epidemiological studies.

-   In Bayesian statistics, every unknown parameter (like a regression
    coefficient, mean, or variance) is treated as a random variable with
    a probability distribution that reflects uncertainty.

-   we then summarize in Bayesian results (posterior mean, credible
    intervals, etc.).

## Model Equation

**Bayesian Logistic Regression model**

$$ \text{logit}(P(Y_i=1)) = \beta_0 + \beta_1 \cdot Age_i + \beta_2 \cdot BMI_i + \beta_3 \cdot Race_i + \beta_4 \cdot Gender_i $$

**Linear Regression equation:**

$$ y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i $$

**Diagnostics performed before Bayesian Regression**

(1) Modeling assumption check
(2) Correlation matrix, Cook's distance, influential points
(3) Model plot

```{r}
#| label: modeling-Survey-weighted complete-case 

# Modeling

library(broom)
library(mice)
library(brms)
library(posterior)
library(bayesplot)
library(knitr)

# --- Guardrails for modeling ---
n_outcome <- sum(!is.na(adult$diabetes_dx))
if (n_outcome == 0) stop("Too few non-missing outcomes for modeling. n = 0")

# Ensure factors and >=2 observed levels among complete outcomes
adult <- adult %>%
  dplyr::mutate(
    sex  = if (!is.factor(sex))  factor(sex)  else sex,
    race = if (!is.factor(race)) factor(race) else race
  )

if (nlevels(droplevels(adult$sex[!is.na(adult$diabetes_dx)]))  < 2)
  stop("sex has <2 observed levels after filtering; check data availability.")
if (nlevels(droplevels(adult$race[!is.na(adult$diabetes_dx)])) < 2)
  stop("race has <2 observed levels after filtering; check Data Prep.")

# ------------------------- 1) Survey-weighted complete-case -------------------------
# Build a logical filter on the original adult data (same length as design$data)
keep_cc <- with(
  adult,
  !is.na(diabetes_dx) & !is.na(age_c) & !is.na(bmi_c) &
  !is.na(sex) & !is.na(race)
)

# Subset the survey design using the logical vector (same length as original)
des_cc <- subset(nhanes_design_adult, keep_cc)

# Corresponding complete-case data (optional)
cc <- adult[keep_cc, ] |> droplevels()
cat("\nComplete-case N for survey-weighted model:", nrow(cc), "\n")

print(table(cc$race))
print(table(cc$diabetes_dx))
print(table(cc$sex))

form_cc <- diabetes_dx ~ age_c + bmi_c + sex + race
svy_fit <- survey::svyglm(formula = form_cc, design = des_cc, family = quasibinomial())
summary(svy_fit)

plot(residuals(svy_fit, type='deviance'))


# Survey-weighted OR table (no intercept)
svy_or <- broom::tidy(svy_fit, conf.int = TRUE) %>%
  dplyr::mutate(OR = exp(estimate), LCL = exp(conf.low), UCL = exp(conf.high)) %>%
  dplyr::select(term, OR, LCL, UCL, p.value) %>%
  dplyr::filter(term != "(Intercept)")
knitr::kable(svy_or, caption = "Survey-weighted odds ratios (per 1 SD)")

```

The residual plot looks okay and does not show any pattern.

```{r}
#| label: MICE
#| include: false
 
# ------------------------- 2) Multiple Imputation (predictors only) 
mi_dat <- adult %>%
  dplyr::select(diabetes_dx, age, bmi, sex, race, WTMEC2YR, SDMVPSU, SDMVSTRA)

meth <- mice::make.method(mi_dat)
pred <- mice::make.predictorMatrix(mi_dat)

# Do not impute outcome
meth["diabetes_dx"] <- ""
pred["diabetes_dx", ] <- 0
pred[,"diabetes_dx"] <- 1

# Imputation methods
meth["age"]  <- "norm"
meth["bmi"]  <- "pmm"
meth["sex"]  <- "polyreg"
meth["race"] <- "polyreg"

# Survey design vars as auxiliaries only
meth[c("WTMEC2YR","SDMVPSU","SDMVSTRA")] <- ""
pred[, c("WTMEC2YR","SDMVPSU","SDMVSTRA")] <- 1

imp <- mice::mice(mi_dat, m = 5, method = meth, predictorMatrix = pred, seed = 123)


```

```{r}
#| label: fit mutated model
#| echo: true
fit_mi <- with(imp, {
  age_c <- as.numeric(scale(age))
  bmi_c <- as.numeric(scale(bmi))
  glm(diabetes_dx ~ age_c + bmi_c + sex + race, family = binomial())
})
pool_mi <- pool(fit_mi)
summary(pool_mi)

## table 

mi_or <- summary(pool_mi, conf.int = TRUE, exponentiate = TRUE) %>%
  dplyr::rename(
    term = term, OR = estimate, LCL = `2.5 %`, UCL = `97.5 %`, p.value = p.value
  ) %>%
  dplyr::filter(term != "(Intercept)")
knitr::kable(mi_or, caption = "MI pooled odds ratios (per 1 SD)")



```

```{r}
#| label: Bayesian model and summary
library(gt)

# 3) Bayesian Logistic Regression (formula weights) 
adult_imp1 <- complete(imp, 1) %>%
  dplyr::mutate(
    age_c  = as.numeric(scale(age)),
    bmi_c  = as.numeric(scale(bmi)),
    wt_norm = WTMEC2YR / mean(WTMEC2YR, na.rm = TRUE),
    # ensure factor refs match survey/MICE:
    race = forcats::fct_relevel(race, "NH White"),
    sex  = forcats::fct_relevel(sex,  "Male")
  ) %>%
  dplyr::filter(!is.na(diabetes_dx), !is.na(age_c), !is.na(bmi_c),
                !is.na(sex), !is.na(race)) %>%
  droplevels()

stopifnot(all(is.finite(adult_imp1$wt_norm)))

priors <- c(
  set_prior("normal(0, 2.5)", class = "b"),
  set_prior("student_t(3, 0, 10)", class = "Intercept")
)

bayes_fit <- brm(
  formula = diabetes_dx | weights(wt_norm) ~ age_c + bmi_c + sex + race,
  data    = adult_imp1,
  family  = bernoulli(link = "logit"),
  prior   = priors,
  chains  = 4, iter = 2000, seed = 123,
  control = list(adapt_delta = 0.95),
  refresh = 0   # quiet Stan output
)

summary(bayes_fit)

prior_summary(bayes_fit)



```
```{r}
#| label: Visualization prior, bmi, age
###
library(ggplot2)

# Example: priors for two coefficients (age and bmi)
prior_draws <- tibble(
  term = rep(c("Age (per 1 SD)", "BMI (per 1 SD)"), each = 4000),
  value = c(rnorm(4000, 0, 2.5), rnorm(4000, 0, 2.5))
)

ggplot(prior_draws, aes(x = value, fill = term)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Prior Distributions for Coefficients",
       x = "Coefficient Value", y = "Density") +
  scale_fill_manual(values = c("skyblue", "orange"))

# Diabetes vs BMI

library(ggplot2)

# Create the plot
p3 <- ggplot(adult_imp1, aes(x = factor(diabetes_dx), y = bmi, fill = factor(diabetes_dx))) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No Diabetes", "1" = "Diabetes")) +
  labs(
    x = "Diabetes Diagnosis",
    y = "BMI",
    title = "BMI Distribution by Diabetes Status"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Save the plot as an image file (PNG)
ggsave("BMI_Distribution_by_Diabetes_Status.png", plot = p3, width = 7, height = 5, dpi = 300)

p3

# logistic regression curve
p4 <- ggplot(adult_imp1, aes(x = bmi, y = diabetes_dx)) +
  geom_point(aes(y = diabetes_dx), alpha = 0.2, position = position_jitter(height = 0.02)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = TRUE, color = "blue") +
  labs(
    x = "BMI",
    y = "Probability of Diabetes",
    title = "Predicted Probability of Diabetes vs BMI"
  ) +
  theme_minimal()

ggsave("Predicted Probability of Diabetes vs BMI.png", plot = p4, width = 7, height = 5, dpi = 300)

p4
```

Once we get posterior draws, we study Summary stats Mean, median, 95%
credible intervals summary(bayes_fit) or posterior_summary(bayes_fit) -
Visualization Distribution shape mcmc_hist(posterior, pars = c("b_age"))
or mcmc_areas() - Pairwise plots Correlations between parameters
mcmc_pairs(posterior) - Posterior predictive checks Compare model
predictions vs observed data pp_check(bayes_fit) - Model comparison
Using LOO or WAIC loo(bayes_fit) or waic(bayes_fit)

Assumptions for Bayesian logistic regression - posterior check - plots
for linearity - mcmc trace plots for convergence - bayes_R2 for model
fit

```{r}
#| label: Assumptions (Bayesian)

summary(bayes_fit)
p5 <- plot(bayes_fit)   # Posterior distributions
p5               

p6 <- pp_check(bayes_fit)      # Posterior predictive checks
mcmc_trace(bayes_fit)    # Convergence (optional)
bayes_R2(bayes_fit)      # Model fit
    # Leave-one-out cross-validation

ggsave("bayes_fit_plot.png", plot = p5, width = 8, height = 6, dpi = 300)

p6
ggsave("pp_check.png", plot = p6, width = 8, height = 6, dpi = 300)




```

Below are reported odds ratio from the posterior predicted values and
the Bayesian regression summary

```{r}
#| label: Posterior ORs and tables
#| echo: true
#| 
# Posterior ORs (drop intercept, clean labels)

bayes_or <- posterior_summary(bayes_fit, pars = "^b_") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("raw") %>%
  dplyr::mutate(
    term = gsub("^b_", "", raw),
    term = gsub("race", "race:", term),
    term = gsub("sex",  "sex:",  term),
    term = gsub("OtherDMulti", "Other/Multi", term),
    term = gsub("OtherHispanic", "Other Hispanic", term),
    OR   = exp(Estimate),
    LCL  = exp(Q2.5),
    UCL  = exp(Q97.5)
  ) %>%
  dplyr::select(term, OR, LCL, UCL) %>%
  dplyr::filter(term != "Intercept")

knitr::kable(
  bayes_or %>%
    dplyr::mutate(dplyr::across(c(OR,LCL,UCL), ~round(.x, 2))),
  digits = 2,
  caption = "Bayesian posterior odds ratios (95% CrI) — reference: NH White (race), Male (sex)"
)

```

```{r}
# Combined table

if (!dir.exists("outputs")) dir.create("outputs", recursive = TRUE)
saveRDS(svy_fit,   "outputs/svy_fit.rds")
saveRDS(pool_mi,   "outputs/pool_mi.rds")
saveRDS(bayes_fit, "outputs/bayes_fit.rds")
saveRDS(svy_or,    "outputs/survey_OR_table.rds")
saveRDS(mi_or,     "outputs/mi_OR_table.rds")
saveRDS(bayes_or,  "outputs/bayes_OR_table.rds")
```




```{r}
#| label: Model results OR
# Results

# ---- Build compact results table (BMI & Age only) ----
library(dplyr); 
library(tidyr); 
library(knitr); 
library(stringr)

# pretty "OR (LCL–UCL)" string

fmt_or <- function(or, lcl, ucl, digits = 2) {
  paste0(
    formatC(or,  format = "f", digits = digits), " (",
    formatC(lcl, format = "f", digits = digits), "–",
    formatC(ucl, format = "f", digits = digits), ")"
  )
}

# guardrails: require these to exist from Modeling
stopifnot(exists("svy_or"), exists("mi_or"), exists("bayes_or"))
for (nm in c("svy_or","mi_or","bayes_or")) {
  if (!all(c("term","OR","LCL","UCL") %in% names(get(nm)))) {
    stop(nm, " must have columns: term, OR, LCL, UCL")
  }
}

svy_tbl   <- svy_or   %>% mutate(Model = "Survey-weighted MLE")
mi_tbl    <- mi_or    %>% mutate(Model = "MICE pooled")
bayes_tbl <- bayes_or %>% mutate(Model = "Bayesian")

all_tbl <- bind_rows(svy_tbl, mi_tbl, bayes_tbl) %>%
  mutate(term = case_when(
    str_detect(term, "bmi_c|\\bBMI\\b") ~ "BMI (per 1 SD)",
    str_detect(term, "age_c|\\bAge\\b") ~ "Age (per 1 SD)",
    TRUE ~ term
  )) %>%
  filter(term %in% c("BMI (per 1 SD)", "Age (per 1 SD)")) %>%
  mutate(OR_CI = fmt_or(OR, LCL, UCL, digits = 2)) %>%
  select(Model, term, OR_CI) %>%
  arrange(
    factor(Model, levels = c("Survey-weighted MLE","MICE pooled","Bayesian")),
    factor(term,  levels = c("BMI (per 1 SD)","Age (per 1 SD)"))
  )

res_wide <- all_tbl %>%
  pivot_wider(names_from = term, values_from = OR_CI) %>%
  rename(
    `BMI (per 1 SD) OR (95% CI)` = `BMI (per 1 SD)`,
    `Age (per 1 SD) OR (95% CI)` = `Age (per 1 SD)`
  )

kable(
  res_wide,
  align = c("l","c","c"),
  caption = "Odds ratios (per 1 SD) with 95% CIs across models"
)


```

```{r}
#| label: post visualization
# Posterior predictive draws

#Posterior predictive checks (binary outcome)
pp_samples <- posterior_predict(bayes_fit, ndraws = 500)  # 500 draws

# Check dimensions
dim(pp_samples)  # rows = draws, cols = observations

# Plot overlay of observed vs predicted counts (duplicate image)
ppc_dens_overlay(y = adult_imp1$diabetes_dx, yrep = pp_samples[1:50, ]) +
  labs(title = "Posterior Predictive Check: Density Overlay") +
  theme_minimal()

ppc_bars(y = adult_imp1$diabetes_dx, yrep = pp_samples[1:50, ])


# Alternative PP plots (histogram / barplot) for binary outcome (bar chart preferred being discrete outcome)

p7 <- ppc_bars(y = adult_imp1$diabetes_dx, yrep = pp_samples[1:50, ]) +
  labs(title = "Posterior Predictive Check: Barplot of Counts") +
  theme_minimal()

ggsave("ppc_bars.png", plot = p7, width = 8, height = 6, dpi = 300)
p7

#PP check for proportions (useful for binary) # mean comparison
## to check if the simulated means match the observed mean

## mean
p8 <- ppc_stat(y = adult_imp1$diabetes_dx, yrep = pp_samples[1:100, ], stat = "mean") +
  labs(title = "Posterior Predictive Check: Mean of Replicates") +
  theme_minimal()
ggsave("ppc_stat_mean.png", plot = p8, width = 8, height = 6, dpi = 300)


p8 

## sd
p9 <- ppc_stat(y = adult_imp1$diabetes_dx, yrep = pp_samples[1:100, ], stat = "sd") +
  labs(title = "PPC: Standard Deviation of Replicates") +
  theme_minimal()
ggsave("ppc_stat_sd.png", plot = p9, width = 8, height = 6, dpi = 300)

p9

# PP checks with bayesplot options
color_scheme_set("blue")
p10 <- ppc_scatter_avg(y = adult_imp1$diabetes_dx, yrep = pp_samples[1:100, ]) +
  labs(title = "Observed vs Predicted (Avg) Posterior Predictive")


ggsave("ppc_scatter_avg.png", plot = p10, width = 8, height = 6, dpi = 300)
p10

```

```{r}
#| label: mcmc, post vs obs


library(brms)
library(dplyr)

# Posterior summary

post_sum <- posterior_summary(bayes_fit)
colnames(post_sum)


library(posterior)
library(bayesplot)

# Extract posterior draws as a draws_df # simulate posterior outcomes
post <- as_draws_df(bayes_fit)

# Check parameter names
colnames(post)


# Density overlay for age and bmi
p11 <- mcmc_areas(post, pars = c( "b_age_c","b_bmi_c","b_sexFemale","b_raceMexicanAmerican", "b_raceOtherHispanic","b_raceNHBlack","b_raceOtherDMulti" ))

ggsave("mcmc_areas.png", plot = p11, width = 8, height = 6, dpi = 300)
p11


###

predicted <- fitted(bayes_fit, summary = TRUE)
observed <- adult_imp1[, c("bmi", "age")]

# Plot for **bmi** (obs vs pred)

library(ggplot2)
p12 <- ggplot(data = NULL, aes(x = observed$bmi, y = predicted[, "Estimate"])) +
  geom_point() +
  geom_errorbar(aes(ymin = predicted[, "Q2.5"], ymax = predicted[, "Q97.5"])) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  xlab("Observed bmi") + ylab("Predicted bmi")

ggsave("plot_bmi_obs_pred.png", plot = p12, width = 8, height = 6, dpi = 300)
p12



```
```{r}
#| label: pred vs obs


# Combine observed and predicted into one data frame
plot_data <- adult_imp1 %>%
  mutate(
    predicted_bmi = predicted[, "Estimate"],
    lower_ci = predicted[, "Q2.5"],
    upper_ci = predicted[, "Q97.5"],
    obs_index = 1:nrow(adult_imp1)  # index for x-axis
  )

# Line plot
p13 <- ggplot(plot_data, aes(x = obs_index)) +
  geom_line(aes(y = bmi, color = "Observed")) +               # observed BMI
  geom_line(aes(y = predicted_bmi, color = "Predicted")) +   # predicted BMI
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.2) +  # uncertainty
  labs(x = "Observation", y = "BMI", color = "Legend") +
  theme_minimal()

ggsave("Line plot_bmi_obs_pred.png", plot = p13, width = 8, height = 6, dpi = 300)
p13

###
summary(adult_imp1$bmi)
summary(plot_data$bmi_c)

```



```{r}
#| label: Pred vs obs age and bmi


prior_summary(bayes_fit)
prior_draws <- tibble(
  term = rep(c("BMI (per 1 SD)", "Age (per 1 SD)"), each = 4000),
  estimate = c(rnorm(4000, 0, 1), rnorm(4000, 0, 1)),
  type = "Prior"
)

summary(prior_draws)


post
head(post)
names(prior_draws)

library(brms)
library(tidyr)

# Extract posterior draws
post <- as_draws_df(bayes_fit) %>%      # bayes_fit = your brms model
  select(b_bmi_c, b_age_c) %>%               # select your coefficient columns
  pivot_longer(
    everything(),
    names_to = "term",
    values_to = "estimate"
  ) %>%
  mutate(
    term = case_when(
      term == "b_bmi_c" ~ "BMI (per 1 SD)",
      term == "b_age_c" ~ "Age (per 1 SD)"
    ),
    type = "Posterior"
  )


## visualization of prior and predicted draws
combined_draws <- bind_rows(prior_draws, post) 

library(ggplot2)

p14 <- ggplot(combined_draws, aes(x = estimate, fill = type)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ term, scales = "free", ncol = 2) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Prior vs Posterior Distributions",
    x = "Coefficient estimate",
    y = "Density",
    fill = ""
  )
p14

ggsave("Prior vs Posterior Distributions_bmi_age.png", plot = p14, width = 8, height = 6, dpi = 300)


#### Compute proportion of diabetes=1 for each draw
pp_proportion <- rowMeans(pp_samples)  # proportion of 1's in each posterior draw


# Summary of posterior proportions
summary(pp_proportion)

# Optional: visualize the posterior probability distribution
pp_proportion_df <- tibble(proportion = pp_proportion)

p15 <- ggplot(pp_proportion_df, aes(x = proportion)) +
  geom_histogram(binwidth = 0.01, fill = "skyblue", color = "black") +
  labs(
    title = "Posterior Distribution of Proportion of Diabetes = 1",
    x = "Proportion of Diabetes = 1",
    y = "Frequency"
  ) +
  theme_minimal()
p15

ggsave("Posterior Distribution of Proportion of Diabetes = 1.png", plot = p15, width = 8, height = 6, dpi = 300)


```
```{r}
#| label: DM prediction vs prevalence

####

library(tidyverse)

# Posterior predicted proportion vector
# pp_proportion <- rowMeans(pp_samples)  # if not already done

known_prev <- 0.089   # NHANES prevalence

# Posterior summary
posterior_mean <- mean(pp_proportion)
posterior_ci <- quantile(pp_proportion, c(0.025, 0.975))  # 95% credible interval

# Create a data frame for plotting
pp_df <- tibble(proportion = pp_proportion)

# Plot
p16 <- ggplot(pp_df, aes(x = proportion)) +
  geom_histogram(binwidth = 0.005, fill = "skyblue", color = "black") +
  geom_vline(xintercept = known_prev, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = posterior_mean, color = "blue", linetype = "solid", size = 1) +
  geom_rect(aes(xmin = posterior_ci[1], xmax = posterior_ci[2], ymin = 0, ymax = Inf),
            fill = "blue", alpha = 0.1, inherit.aes = FALSE) +
  labs(
    title = "Posterior Predicted Diabetes Proportion vs NHANES Prevalence",
    subtitle = paste0("Red dashed = NHANES prevalence (", known_prev, 
                      "), Blue solid = Posterior mean (", round(posterior_mean,3), ")"),
    x = "Proportion of Diabetes = 1",
    y = "Frequency"
  ) +
  theme_minimal()

ggsave("Posterior Distribution of Proportion of Diabetes = 1.png", plot = p16, width = 8, height = 6, dpi = 300)
p16

```





```{r}
#| label: posterior DM prediction vs population prevalence
library(dplyr)

# Posterior predicted proportion
posterior_mean <- mean(pp_proportion)
posterior_ci <- quantile(pp_proportion, c(0.025, 0.975))  # 95% credible interval

# NHANES prevalence with SE from survey::svymean
# Suppose you already have:
# svymean(~diabetes_dx, nhanes_design_adult, na.rm = TRUE)
known_prev <- 0.089        # Mean prevalence
known_se   <- 0.0048       # Standard error from survey

# Calculate 95% confidence interval
known_ci <- c(
  known_prev - 1.96 * known_se,
  known_prev + 1.96 * known_se
)

# Print results
data.frame(
  Type = c("Posterior Prediction", "NHANES Prevalence"),
  Mean = c(posterior_mean, known_prev),
  Lower_95 = c(posterior_ci[1], known_ci[1]),
  Upper_95 = c(posterior_ci[2], known_ci[2])
)


####
library(ggplot2)
library(dplyr)

# Create a data frame for plotting
ci_df <- data.frame(
  Type = c("Posterior Prediction", "NHANES Prevalence"),
  Mean = c(0.1096674, 0.089),
  Lower_95 = c(0.09772443, 0.079592),
  Upper_95 = c(0.1210658, 0.098408)
)

# Plot
p17 <- ggplot(ci_df, aes(x = Type, y = Mean, color = Type)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = Lower_95, ymax = Upper_95), width = 0.2) +
  ylim(0, max(ci_df$Upper_95) + 0.02) +
  labs(
    title = "Comparison of Posterior Predicted Diabetes Proportion vs NHANES Prevalence",
    y = "Proportion of Diabetes",
    x = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

ggsave("Population Prevalence vs Proportion of Diabetes = 1.png", plot = p17, width = 8, height = 6, dpi = 300)
p17

```

# Results

1.  **Multiple Linear Regression**

    Model**:**
    `svyglm(formula = form_cc, design = des_cc, family = quasibinomial())`

All predictors are significant: **age (p \< 0.001)** and **BMI (p \<
0.001)** show strong positive associations with the outcome, while
**being female (p = 0.0004)** is negatively associated. Other
significant associations include **raceMexican American (p = 0.0008)**,
**raceOther Hispanic (p = 0.0087)**, **raceNH Black (p = 0.0117)**, and
**raceOther/Multi (p = 0.0014)**.

2.  **MICE**

    All predictors are statistically significant.

    Positive associations: age (p \< 0.001), BMI (p \< 0.001), and all
    race categories compared to reference.

    Negative association: being female (p \< 0.001)

<!-- -->

3.  **Bayesian Regression**

    Sampling**:** NUTS (4 chains, 2000 iterations each; 1000 warmup,
    4000 post-warmup draws)

    Convergence & Diagnostics

    -   Rhat = 1.00 for all parameters → excellent convergence

    -   Bulk_ESS / Tail_ESS: Large values (\>2000) → high effective
        sample sizes, reliable posterior estimates.

        Interpretation

    -   Strong predictors: Age and BMI are strongly positively
        associated with diabetes risk.

    -   Sex effect: Females have a lower probability of diabetes
        compared to males

    -   R² = 0.13 shows 13% of the variance in the outcome (diabetes_dx)
        is explained by your predictors (age, BMI, sex, race), 95%
        Credible Interval: 0.106–0.156, indicates that, given your model
        and data, the true proportion of variance explained is plausibly
        between 10.6% and 15.6% and shows uncertainty in the explained
        variance, which is natural for probabilistic models.

    -   Est.Error = 0.0127, reflects the standard error of the R²
        estimate across posterior samples. The small SE indicates that
        the R² estimate is fairly precise.

    <!-- -->

    -   Race/ethnicity: Mexican American, NH Black, and “Other/Diverse”
        groups have higher odds of diabetes. Other Hispanic group has a
        less certain effect.

    <!-- -->

    -   age_c-Each 1-unit increase in centered age increases the
        log-odds of diabetes by 1.09. Strong positive effect.

    -   bmi_c-Higher BMI is associated with higher diabetes risk.
        sexFemale-Females have lower odds of diabetes compared to males.

    -   raceMexicanAmerican-Higher odds of diabetes vs. reference race
        (likely NH White)

    -   raceOtherHispanic-Slightly higher odds vs reference, but
        interval crosses zero → uncertain effect.

    -   raceNHBlack-Significantly higher odds of diabetes compared to
        reference. raceOtherDMulti-Higher odds of diabetes vs reference
        group.

**Posterior distribution** of all parameters in the model. (1)Density
plot of posterior samples each parameter (e.g., intercept, slope) into a
smoothed density curve, showing most of the posterior probability mass
lies for best estimates and uncertainty.

**Posterior Predictive Distribution** - generated from posterior
predictive draws: 𝑦rep∼𝑝(𝑦new∣𝜃)yrep​∼p(ynew​∣θ) simulate the data given
posterior parameter estimates.Posterior predictive checks (PPC) compare
these simulations to real data to assess model fit.

**Incorporating Uncertainty** two sources of uncertainty: Parameter
uncertainty: captured in the posterior distributions Predictive
uncertainty: captured in posterior predictive draws

Combining the two provide credible intervals for predictions, not just
point predictions and specifies - Given the BMI, the probability of
diabetes is likely between 40–55%.”

**Comparing Models**

-   All three models (survey-weighted MLE, multiple imputation,
    Bayesian) agree closely on the direction and magnitude of the
    effects of BMI and age.

-   Age is a stronger predictor than BMI, nearly tripling the odds per 1
    SD.

-   BMI significantly increases diabetes risk (\~1.7–1.9× per 1 SD).

-   Differences between models are minor, indicating robust and reliable
    findings despite missing data or modeling approach.
    
## **Diabetes Predicted proportion vs population prevalence**
    
 - The posterior predicted proportion is slightly higher than the observed NHANES prevalence (10.97% vs 8.9%).
 - Intervals overlap slightly, but the posterior tends to overestimate diabetes compared to NHANES.

The difference could be due to:

- Model assumptions (logistic link, priors)
- Predictor effects (BMI, age, sex, race)
- Sample characteristics vs population weighting in NHANES

The results find our model is reasonable but slightly conservative (predicting higher risk) relative to the observed population prevalence.

# Conclusion

-   Across multiple modeling approaches—survey-weighted maximum
    likelihood, multiple imputation, and Bayesian regression—both age
    and BMI were consistently strong predictors of diabetes. 
    Eachstandard deviation increase in age nearly tripled the odds of
    diabetes, while a similar increase in BMI elevated the odds by
    approximately 1.7–1.9 times.
    The consistency of these results across
    models highlights the robustness of the associations and underscores
    the importance of age and BMI as key risk factors for diabetes in
    this population.

    Effect stability: point estimates in rhe Bayesian model’s closely
  aligned with those from the frequentist, indicating that prior
  regularization stabilized the estimates in the presence of modest
  missingness.

  Uncertainty quantification: Bayesian credible intervals of odds ration
  were slightly narrower yet overlapped the frequentist confidence
intervals, suggest comparable inferential precision while offering
improved interpretability.

  Design considerations: \# Survey-weighted MLE (Maximum Likelihood
Estimator) - incorporates each observation weighted according to its
survey weight. - provide estimates that reflect the population-level
parameters, not just the sample- produces population-representative
estimates. \# Bayesian model with normalized weights- - instead of fully
modeling the survey design, it used normalized sampling weights as
importance weights - the scaled weights that sum to the sample size
approximates the effect of survey weights, but does not fully account
for: Stratification, clustering, design-based variance adjustments. -
Bayesian inference treats the weighted likelihood as from a regular
model, ignoring some survey design features.

# Discussions

The use of multiple imputation allowed for robust analysis despite
missing data, increasing power and reducing bias. Comparison of
frequentist and Bayesian models demonstrated consistency in significant
predictors, while Bayesian approaches provided the advantage of
posterior distributions and probabilistic interpretation. The =

Across all models, both age and BMI emerged as strong and consistent
predictors of diabetes. The consistency across modeling approaches
strengthens the validity of these findings Multiple imputation accounted
for potential biases due to missing data, and Bayesian modeling provided
robust credible intervals that closely matched frequentist estimates.
align with previous epidemiological research indicating that increasing
age and higher BMI are among the most important determinants of type 2
diabetes risk.Cumulative exposure to metabolic and lifestyle risk
factors over time, and the role of excess adiposity and insulin related
effects account for diabetes.

Survey weighted dataset strenghthens ensuring population
representativeness, multiple imputation to handle missing data, and
rigorous Bayesian estimation provided high effective sample sizes and R̂
≈ 1.00 across parameters confirmed excellent model convergence. Bayesian
logistic regression provided inference statistically consistent and
interpretable achieving the aim of this study. In future research
hierarchical model using NHANES cycles or adding variables (lab tests)
could assess nonlinear effects of metabolic risk factors.

# Limitations

Our study was a cross-sectional study design - precludes potential
residual confounding from unmeasured factors such as diet, physical
activity, and genetic predisposition

# Implications

-   age and BMI as robust and independent predictors of diabetes,
    underscore the importance of early targeted interventions in
    mitigating diabetes risk.
-   Longitudinal studies and combining other statistical analytical
    methods with Bayesian can further enhance and provide better
    informed precision prevention strategies.

## References
