
---
title: "Bayesian Logistic Regression Application in Diabetes Probability Prediction"
subtitle: "Capstone Project Presentation"
author: "Namita Mishra (Advisor: Dr. Ashraf Cohen)"
date: '`r Sys.Date()`'
format: revealjs
course: Capstone Projects in Data Science
bibliography: references.bib
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---



## Type 2 Diabetes (T2D), a public health concern worldwide

- Body doesn’t make enough insulin or is not used properly → rise in blood sugar
- Affects organs and can lead to multi-organ failure
- Understanding risk factors (age, BMI, genetics) allows early intervention

---

## Statistical Problem

- Traditional statistical approaches fail to analyze complex relationships or uncertainty present in healthcare data
- Missing data in survey-collected datasets reduces sample size and may bias estimates

---

## Study Goal

- Address statistical challenges where traditional methods fail
- Compare Frequentist and Bayesian methods on NHANES 2013–2014
- Identify associations between risk factors and diabetes (dichotomous outcome)
- Demonstrate model performance and insights for healthcare data analysis

---

## Methods

- Exploratory Data Analysis (EDA) of weighted NHANES dataset revealed missing values
- Multivariate linear regression on complete cases resulted in reduced sample size
- Multiple Imputation by Chained Equations (MICE) was applied to handle missing data
- Bayesian regression was conducted on imputed data after normalizing age and BMI

---

## Bayesian Regression Principles

- Priors stabilize estimates and reduce extreme values
- Handles uncertainty fully through posterior distributions
- Incorporates prior knowledge for robust inference
- Works effectively with imputed datasets

---

## Data Source

- NHANES 2013–2014 (CDC)  
- Files merged: demographics (DEMO_H), exam (BMX_H), questionnaire (DIQ_H)  
- Cleaned dataset: n = 10,175 observations, 10 variables

| Dataset | Key variables     | Dimensions |
|---------|------------------|------------|
| demo1   | age, race, gender | 10175 × 47 |
| exam1   | BMI               | 9813 × 224 |
| quest1  | Diabetes          | 9813 × 953 |

---

## Data Exploration {.smaller}

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(viridis)
library(bayesplot)
library(dslabs)
theme_set(theme_minimal(base_size = 13))
custom_colors <- viridis(3, option = "D")

```
* Small sample for certain BMI subgroups (pre-diabetic n=132/9813)
* Only 14 complete cases across all variables → imbalance
* Survey-weighted proportions approximate US population (Male:Female ~48.9%:51%)
* Predictor correlations are hierarchical and complex → suitable for Bayesian modeling

---

## Our Question

* Can Bayesian logistic regression provide **more stable** and **transparent** inference than classical MLE for diabetes outcomes in NHANES 2013–2014?

---

## Data Pipeline (Reproducible)

```{r data-pipeline, echo=FALSE}
# Load merged NHANES data
merged_data <- readRDS("data/merged_2013_2014.rds")
knitr::kable(head(merged_data))
```

* Adults >20 years filtered: n = 5,769
* Standardized predictors: age_c, bmi_c
* Outcome adjusted for pregnancy when female
* NH White set as reference level for race

```

---




 


